Assessing linkages between watershed nutrient loading and water quality in a subtropical estuary with semiparametric models.

### 2023-05-20 Schramm-2023-05-AS/Schramm-2023-05-AS.tex
init springer nature tex submission for AS

%  LaTeX support: latex@mdpi.com
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

%=================================================================
% pandoc conditionals added to preserve backwards compatibility with previous versions of rticles
\documentclass[water,article,submit,oneauthor]{Definitions/mdpi}

% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'. The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

%% Some pieces required from the pandoc template
\setlist[itemize]{leftmargin=*,labelsep=5.8mm}
\setlist[enumerate]{leftmargin=*,labelsep=4.9mm}

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, adolescents, aerobiology, aerospace, agriculture, agriengineering, agrochemicals, agronomy, ai, air, algorithms, allergies, alloys, analytica, analytics, anatomia, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arm, arthropoda, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coasts, coatings, colloids, colorants, commodities, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryptography, crystals, csmf, ctn, curroncol, cyber, dairy, data, ddc, dentistry, dermato, dermatopathology, designs, devices, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, entomology, entropy, environments, environsciproc, epidemiologia, epigenomes, est, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, foundations, fractalfract, fuels, future, futureinternet, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, grasses, gucdd, hazardousmatters, healthcare, hearts, hemato, hematolrep, heritage, higheredu, highthroughput, histories, horticulturae, hospitals, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijpb, ijtm, ijtpp, ime, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jcto, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jvd, jzbg, kidneydial, kinasesphosphatases, knowledge, land, languages, laws, life, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, microplastics, minerals, mining, modelling, molbank, molecules, mps, msf, mti, muscles, nanoenergyadv, nanomanufacturing,\gdef\@continuouspages{yes}} nanomaterials, ncrna, ndt, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, %%nri, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, %oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacoepidemiology,\gdef\@ISSN{2813-0618}\gdef\@continuous pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, platforms, pollutants, polymers, polysaccharides, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, publications, quantumrep, quaternary, qubs, radiation, reactions, receptors, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, ruminants, safety, sci, scipharm, sclerosis, seeds, sensors, separations, sexes, signals, sinusitis, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, spectroscj, sports, standards, stats, std, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, targets, taxonomy, technologies, telecom, test, textiles, thalassrep, thermo, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, virtualworlds, viruses, vision, waste, water, wem, wevj, wind, women, world, youth, zoonoticdis
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal. Changing "submit" to "accept" before posting will remove line numbers.

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by:
% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. Remove "pdftex" for (1) compiling with LaTeX & dvi2pdf (if eps figures are used) or for (2) compiling with XeLaTeX.

%=================================================================
% MDPI internal commands - do not modify
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2023}
\copyrightyear{2023}
%\externaleditor{Academic Editor: Firstname Lastname}
\datereceived{ }
\daterevised{ } % Comment out if no revised date
\dateaccepted{ }
\datepublished{ }
%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%\pdfoutput=1 % Uncommented for upload to arXiv.org

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}

%=================================================================
% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Assessing linkages between watershed nutrient loading and estuary
water quality in Lavaca Bay, Texas}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Assessing linkages between watershed nutrient loading and
estuary water quality in Lavaca Bay, Texas}

% Author Orchid ID: enter ID or remove command
%\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Michael
Schramm$^{1}$\href{https://orcid.org/0000-0003-1876-6592}
{\orcidicon}}

%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Michael Schramm}

% MDPI internal command: Authors, for citation in the left column
%\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.
\AuthorCitation{Schramm, M.}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Texas A\&M AgriLife Research - Texas Water Resources
Institute 1001 Holleman Dr.~E. College Station, TX
77840-2118; \href{mailto:michael.schramm@ag.tamu.edu}{\nolinkurl{michael.schramm@ag.tamu.edu}}\\
}

% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:michael.schramm@ag.tamu.edu}{\nolinkurl{michael.schramm@ag.tamu.edu}}}

% Current address and/or shared authorship

% The commands \thirdnote{} till \eighthnote{} are available for further notes

% Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Lavaca Bay is a small secondary embayment on the Texas coast
that is displaying early signals of water quality degradation. This
study uses Generalized Additive Models to assesses both watershed
nutrient loading and responses in estuary water quality to nutrient
loading and streamflow. While delivered nutrient loads varied greatly,
flow-normalized nutrient loads indicated limited evidence of changes in
nutrient runoff in the watershed since 2005. As a consequence, there
wasn't strong evidence for an effect of nutrient loading (controlled for
flow) on dissolved oxygen or chlorophyll-\emph{a} concentrations. There
was evidence total phosphorus (TP) loading had a strong positive effect
on TP concentration in the Bay, particularly at sites closest to the
river mouth. Seasonally-adjusted streamflow had strong effects on
nutrient concentration and chlorophyll-\emph{a} concentrations at most
sites. Although temporal trends did not identify significant estuary
water quality degredation in dissolved oxygen or chlorophyll-\emph{a}
concentrations, site specific trends in both organic and inorganic
nitrogen are concerning for the long-term health of the estuary. This
study provides a baseline assessment of the linkages between watershed
nutrient loading and water quality responses in Lavaca Bay. Some of the
data gaps highlighted in the study are being addressed in ongoing
projects to provide improved information for resource managers.}

% Keywords
\keyword{estuary; nutrient loading; water quality; Texas}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data
%\dataset{DOI number or link to the deposited data set if the data set is published separately. If the data set shall be published as a supplement to this paper, this field will be filled by the journal editors. In this case, please submit the data set as a supplement.}
%\datasetlicense{License under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Encyclopedia
%\encyclopediadef{For entry manuscripts only: please provide a brief overview of the entry title instead of an abstract.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Advances in Respiratory Medicine
%\addhighlights{yes}
%\renewcommand{\addhighlights}{%

%\noindent This is an obligatory section in “Advances in Respiratory Medicine”, whose goal is to increase the discoverability and readability of the article via search engines and other scholars. Highlights should not be a copy of the abstract, but a simple text allowing the reader to quickly and simplified find out what the article is about and what can be cited from it. Each of these parts should be devoted up to 2~bullet points.\vspace{3pt}\\
%\textbf{What are the main findings?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}\vspace{3pt}
%\textbf{What is the implication of the main finding?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Like many coastal areas globally, coastal watersheds along the Texas
Gulf coast are facing pressures from increasing population, increases in
point source and non-point source pollution and alterations to
freshwater flows that affect water quality in downstream estuaries
\citep{bricker_effects_2008, kennicuttWaterQualityGulf2017, bugica_water_2020}.
Despite these increasing pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or lower for
exhibiting eutrophic conditions \citep{bricker_effects_2008}. However, a
suite of recent studies indicate that estuary water quality dynamics in
both agriculturally dominated and urban watersheds within Texas are in
fact expressing conditions that are increasingly conducive to algal
blooms and eutrophication
\citep{wetzWaterQualityDynamics2016, wetz_exceptionally_2017, bugica_water_2020, chinPhytoplanktonBiomassCommunity2022}.
With identification of localized areas of estuary water quality concern
along the Texas coast \citep{bugica_water_2020}, localized studies are
being prioritized to better inform management actions.

This project provides an assessment of nutrient loading and water
quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in
the larger Matagorda Bay system located roughly halfway between Houston,
Texas and Corpus Christi, Texas. Lavaca Bay has faced substantial
challenges associated with legacy contamination but general water
quality parameters such as dissolved oxygen (DO), nutrients, and
biological parameters have been well within state water quality
standards. However, long-term declines of benthic fauna abundance,
biomass, and diversity in Lavaca Bay primarily linked to reductions in
freshwater inflows and changes in estuary salinity
\citep{beserespollackLongtermTrendsResponse2011, palmerImpactsDroughtsLow2015, montagnaAssessmentRelationshipFreshwater2020}
are a concern to local stakeholders. \citet{bugica_water_2020}
identified monotonic increases in total phosphorus (TP), orthophosphate,
total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a} at sites within
Lavaca Bay. Although long-term changes in DO concentrations were not
identified, the trends in nutrient concentrations are concerning due to
the role of nitrogen as a limiting factor for primary production in many
Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the ramifications that changes in nitrogen loadings could have for
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\citep{jainTechnicalSupportDocument2021, schramm_lavaca_2018, bertholdDirectMailingEducation2021}.
However, at a statewide scale, these approaches have shown limited
success and emphasize a need for improved efforts at assessing and
linking management actions with downstream water quality to better
identify and replicate effective management actions across the state
\citep{schrammTotalMaximumDaily2022}. The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\citep{wazniakLinkingWaterQuality2007, lloydMethodsDetectingChange2014}.
To provide actionable information for resource managers, water quality
conditions must be evaluated relative to changes in natural
environmental drivers to better understand and manage potential
anthropogenic effects. This study utilizes semiparemetric methods to
develop estimates of delivered and flow-normalized nutrient loads and
assess changes in loads delivered to Lavaca Bay. The study also assesses
the response of water quality parameters in Lavaca Bay over time and in
response to freshwater inflow controlled for seasonality and to
watershed nutrient loads that are controlled for environmentally driven
variation.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{study-area-and-data}{%
\subsection{Study Area and Data}\label{study-area-and-data}}

\begin{figure}
\includegraphics[width=1\linewidth]{Schramm-Manuscript-2023_files/figure-latex/fig1-1.png}
\caption{Map of the Lavaca Bay watershed, location of USGS gages where nutrient loads were calculated, and location of estuary water quality sampling sites.\label{fig:fig1}}
\end{figure}

Lavaca Bay is 190 km\textsuperscript{2} with the majority of freshwater
inflow provided by the Lavaca and Navidad River systems (Figure
\ref{fig:fig1}). The Garcitas-Arenosa, Placedo Creek, and Cox Bay
watersheds provide additional freshwater inflows. The entire watershed
land area is 8,149 km\textsuperscript{2} and primarily rural. Watershed
land cover and land use is 50\% grazed pasture and rangeland, 20\%
cultivated cropland (primarily rows crops such as corn, cotton, and
sorghum), and 5\% suburban/urban. Pasture and rangeland is concentrated
in the Lavaca River watershed, while cultivated crops are generally
located along the eastern tributaries of the Navidad river. The Lavaca
and Navidad River watersheds are a combined 5,966 km\textsuperscript{2},
or approximately 73\% of the entire Lavaca Bay watershed area. Discharge
from the Navidad River is regulated by Lake Texana which has been in
operation since 1980. Lake Texana provides 0.210 km\textsuperscript{3}
of water storage and discharges into the tidal section of the Navidad
River which ultimately joins the tidal section of the Lavaca River 15 km
upstream of the confluence with the Lavaca Bay.

Daily discharges for the Lavaca River (USGS-08164000, Figure
\ref{fig:fig1}) were obtained from the United States Geologic Survey
(USGS) National Water Information System using the \emph{dataRetrieval}
R package \citep{deciccoDataRetrievalPackagesDiscovering2022}. Gaged
daily discharges from the outlet of Lake Texana on the Navidad River
(USGS-0816425) were provided by the Texas Water Development Board (TWDB)
(April 21, 2022 email from R. Neupane, TWDB).

Water quality sample data for the two freshwater and three estuary
locations were obtained from the Texas Commission on Environmental
Quality (TCEQ) Surface Water Quality Monitoring Information System. Data
submitted through the system are required to be collected under Quality
Assurance Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. The QAPP and procedures manuals ensure the consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying lengths of and availability of data. For freshwater locations,
TP from January 2000 through December 2020 and nitrate-nitrogen
(NO\textsubscript{3}) data from January 2005 through December 2020 were
downloaded (Table \ref{tab:fwsummary}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater
sites and deemed insufficient to develop load estimation models
\citep{horowitzEvaluationSedimentRating2003, snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper Lavaca Bay site near the
outlet of the Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site
(TCEQ-13383), and the lower Lavaca Bay site near the mouth of the Bay
(TCEQ-13384). For estuary locations, we obtained data for TP,
Nitrite+Nitrate (NO\emph{\textsubscript{x}}), TKN, chlorophyll-\emph{a},
and DO concentrations from January 2005 through December 2020 (Table
\ref{tab:estuarysummary}).

\begin{table}[H]

\caption{\label{tab:fwsummary}Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020.}
\centering
\begin{tabular}[t]{llrrr}
\toprule
Station ID &   & Mean & SD & N\\
\midrule
USGS-08164000 & TP (mg/L) & \num{0.21} & \num{0.09} & 80\\
 & NO\textsubscript{3} (mg/L) & \num{0.18} & \num{0.24} & 74\\
 & Mean Daily Streamflow (cfs) & \num{332.78} & \num{1667.47} & 7671\\
USGS-08164525 & TP (mg/L) & \num{0.20} & \num{0.08} & 81\\
 & NO\textsubscript{3} (mg/L) & \num{0.29} & \num{0.26} & 62\\
 & Mean Daily Streamflow (cfs) & \num{666.14} & \num{2957.79} & 7671\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]

\caption{\label{tab:estuarysummary}Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.}
\centering
\begin{tabular}[t]{lllll}
\toprule
Station ID &   & Mean & SD & N\\
\midrule
 & TP (mg/L) & 0.11 & 0.05 & 47\\

 & NO\textsubscript{\emph{x}} (mg/L) & 0.07 & 0.15 & 51\\

 & TKN (mg/L) & 0.94 & 0.49 & 45\\

 & Chlorophyll-\emph{a} ($\mu$g/L) & 9.43 & 5.31 & 47\\

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13383} & DO (mg/L) & 7.22 & 1.35 & 55\\
\cmidrule{1-5}
 & TP (mg/L) & 0.08 & 0.03 & 51\\

 & NO\textsubscript{\emph{x}} (mg/L) & 0.06 & 0.08 & 52\\

 & TKN (mg/L) & 0.76 & 0.40 & 48\\

 & Chlorophyll-\emph{a} ($\mu$g/L) & 8.22 & 6.44 & 46\\

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13384} & DO (mg/L) & 7.51 & 1.32 & 54\\
\cmidrule{1-5}
 & TP (mg/L) & 0.13 & 0.06 & 50\\

 & NO\textsubscript{\emph{x}} (mg/L) & 0.09 & 0.13 & 53\\

>>>>

%Version 2.1 April 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[sn-basic,referee,lineno,pdflatex]{sn-jnl}

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

@@

\raggedbottom

@@

\title[Watershed nutrient loading and water quality in a subtropical
estuary]{Assessing linkages between watershed nutrient loading and water
quality in a subtropical estuary with semiparametric models.}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1]{\fnm{Michael} \sur{Schramm} }\email{\href{mailto:michael.schramm@ag.tamu.edu}{\nolinkurl{michael.schramm@ag.tamu.edu}}}

  \affil*[1]{\orgdiv{Texas Water Resources Institute}, \orgname{Texas
A\&M AgriLife Research}, \orgaddress{\city{College
Station}, \country{United
States}, \postcode{77840}, \state{TX}, \street{1001 Holleman Dr.~E.}}}

\abstract{Lavaca Bay is a small secondary embayment on the Texas coast
that is displaying early signals of water quality degradation. This
study applied a semiparametric approach to assess both watershed
nutrient loads and the responses in estuary water quality to nutrient
loading and streamflow. Cross-validation indicated that, despite data
constraints, semiparametric models performed well at nutrient load
prediction. Based on these models, delivered annual nutrient loads
varied substantially from year to year. In contrast, minimal changes in
calculated flow-normalized loads indicate that changes in nutrient
loading were driven by natural variation in precipitation and runoff as
opposed to changes in nonpoint sources. Estuary water quality models did
not identify significant long-term changes within Lavaca Bay for
dissolved oxygen or chlorophyll-\emph{a}. However, site specific
long-term increases in both organic and inorganic nitrogen are
concerning. Further analysis found freshwater inflow was a strong driver
of nutrient and chlorophyll-\emph{a} concentrations within Lavaca Bay
but there was no evidence that changes in nutrient loading explained
variation in dissolved oxygen or chlorophyll-\emph{a} concentrations. In
addition to providing a baseline assessment of watershed nutrient
loading and water quality responses in the Lavaca Bay watershed, this
study provides methodological support for the use of semiparametric
methods in load regression models and estuary assessments.}

\keywords{key, dictionary, word}

\maketitle

\hypertarget{sec1}{%
\section{Introduction}\label{sec1}}

Like many coastal areas globally, the coastal watersheds along the Texas
Gulf coast are facing pressures from increasing population, increases in
point source and non-point source pollution and alterations to
freshwater flows that degrade water quality in downstream estuaries
\citep{bricker_effects_2008, kennicuttWaterQualityGulf2017, bugica_water_2020}.
Despite these escalating pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or low risk for
eutrophic conditions \citep{bricker_effects_2008}. However, a suite of
recent studies indicates that estuary water quality dynamics in both
agriculturally dominated and urban watersheds within Texas are in fact
expressing conditions that are increasingly conducive to algal blooms
and eutrophication
\citep{wetzWaterQualityDynamics2016, wetz_exceptionally_2017, bugica_water_2020, chinPhytoplanktonBiomassCommunity2022}.
With identification of localized areas of estuary water quality concern
along the Texas coast \citep{bugica_water_2020}, localized studies are
being prioritized to better inform management actions.

This project aims to provide an assessment of nutrient loading and water
quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in
the larger Matagorda Bay system located roughly halfway between Houston,
Texas and Corpus Christi, Texas. Lavaca Bay faces substantial challenges
associated with legacy contamination but general water quality
parameters such as dissolved oxygen (DO), nutrients, and biological
parameters have been well within state water quality standards. More
recently identified long-term declines in abundance, biomass, and
diversity of benthic fauna in Lavaca Bay have been linked to reductions
in freshwater inflows and changes in estuary salinity
\citep{beserespollackLongtermTrendsResponse2011, palmerImpactsDroughtsLow2015, montagnaAssessmentRelationshipFreshwater2020}
and are a concern to local stakeholders. Water quality assessments
identified monotonic increases in total phosphorus (TP), orthophosphate,
total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a} at sites within
Lavaca Bay \citep{bugica_water_2020}. Although long-term changes in DO
concentrations were not identified, the trends in nutrient
concentrations are concerning due to the role of nitrogen as a limiting
factor for primary production in many Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the ramifications that changes in nitrogen loadings could have for
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\citep{jainTechnicalSupportDocument2021, schrammLavacaRiverWatershed2018, bertholdDirectMailingEducation2021}.
However, at a statewide scale, these approaches have shown limited
success and emphasize a need for improved efforts at assessing and
linking management actions with downstream water quality to identify and
replicate effective management actions across the state
\citep{schrammTotalMaximumDaily2022}. The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\citep{wazniakLinkingWaterQuality2007, lloydMethodsDetectingChange2014}.
The development and application of statistical methods such as Weighted
Regressions on Time, Discharge and Season
\citep[WRTDS,][]{hirschWeightedRegressionsTime2010} and Generalized
Additive Models \citep[GAMs,][]{woodFastStableRestricted2011} has
provided effective tools for researchers to quantify and communicate
non-linear changes in river and estuary pollutant loadings.

WRTDS calculates a time series of in stream concentrations or loads
(daily, monthly, or annually) and flow-normalized estimates of
concentrations and loads using locally weighted regresssion for unique
combinations of time, discharge, and season. WRTDS has been widely used
to assess and identify trends in riverine nutrients
\citep{oelsner_recent_2019, stackpooleLongTermMississippi2021},
chlorides \citep{stetsIncreasingChlorideRivers2018}, and other
pollutants of concern \citep{shodaWaterqualityTrendsRivers2019}. WRTDS
has also been succesfully adapted to assess trends in estuarine water
quality concentrations \citep{beckFourDecadesWater2018}.

While WRTDS is a statistical approach developed specifically for water
quality applications, GAMs are a broadly applicable statistical method.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\citep{woodFastStableRestricted2011}. Although the underlying parameter
estimation procedure of GAMs is substantially different than WRTDS, both
the functional form and results have been demonstrated to be similar
\citep{beckNumericalQualitativeContrasts2017}. Water quality
applications of GAMs include river and catchment nutrient concentration
and load models
\citep{wangLoadEstimationUncertainties2011, kroonRiverLoadsSuspended2012, kuhnertQuantifyingTotalSuspended2012, robson_prediction_2015-1, hagemannEstimatingNutrientOrganic2016, mcdowell_implications_2021, biagi_novel_2022}.
GAMs can also be used to identify non-linear temporal trends (including
flow-normalized trends) in pollutant concentrations and loads
\citep{beckNumericalQualitativeContrasts2017, murphyGeneralizedAdditiveModel2019}.
Recently GAMs have also been used to link water quality responses in
receiving water bodies to changes in nutrient inputs
\citep{murphyNutrientImprovementsChesapeake2022}.
\citet{beckNumericalQualitativeContrasts2017} provides a substantial
discussion on the differences (and similarities) between GAMs and WRTDS
for water quality applications.

To provide actionable information for resource managers in Lavaca Bay,
water quality conditions must be evaluated relative to changes in
natural environmental drivers to better understand and manage potential
human impacts. This study utilizes GAMs to develop estimates of
delivered and flow-normalized nutrient loads and assess changes in loads
delivered to Lavaca Bay. GAMs were chosen over other regression-based
approached for use in this study due to; (1) the ability to easily
explore and incorporate different model terms; (2) the incorporation of
non-liner smooth functions that do not require explicit a priori
knowledge of the expected shape; and (3) inclusion of a link function
that related the expected value of the response to linear predictors
thus avoiding unneeded data transformations and bias corrections. The
study also assesses the response of water quality parameters in Lavaca
Bay over time and in response to freshwater inflow controlled for
seasonality and to watershed nutrient loads that are controlled for
environmentally driven variation.

\hypertarget{sec2}{%
\section{Methods}\label{sec2}}

\hypertarget{location-and-data}{%
\subsection{Location and Data}\label{location-and-data}}
